Multi-resolution Neural Field Estimation

### 2011-03-08 Multiresolution_framework/ltx/manuscript.tex
davids edits to the discussion

\usepackage{color}
\newcommand{\mike}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\dean}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}} 
\newcommand{\cut}[1]{\textcolor{cyan}{#1}}

@@

\begin{abstract}
%\boldmath
The Integro-difference equation (IDE) is an increasingly popular model of spatio-temporal processes. Here we develop a multi-resolution approximation (MRA) framework for the IDE neural field equations based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximisation (EM) algorithm. A synthetic example is given
to demonstrate the framework.
\end{abstract}
% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% Neural field model, multiresolution approximation (MRA), Expectation Maximization (EM) algorithm, wavelets.
% \end{IEEEkeywords}
% \IEEEpeerreviewmaketitle
\section{Introduction}
\IEEEPARstart{T}{he} human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE), to address this challenge.

@@

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where \mike{a finite element model (FEM)} of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. The current paper is an extension to this recent study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behaviour of the system can be represented simultaneously.

@@

Each $1\times n_{\theta}$ block of the $n_x \times n_x n_{\theta}$ block matrix $\mathbf U$ is given by 

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1 second of data were generated \parham{and the estimation was applied to the final 900 ms, allowing the model's initial transients to die out.} The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  %\ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} 
The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, \parham{This is obtained by examining the average (over time) power of the spatial frequency of the observations} \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). \parham{It is assumed that the support of the kernel is unknown to the estimator.} The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 1.93$. (b) $j=1$, $n_x=33$, $RMSE = 1.39$. (c) $j=2$, $n_x=65$, $RMSE = 0.90$. (d) $j=3$, $n_x=131$, $RMSE = 0.78$. (e) $j=4$, $n_x=263$, $RMSE=0.78$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig2.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The mean kernel estimates over 200 realizations and \mike{confidence interval} are shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution \mike{FEM} representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

 \cut{The MRAIDE approach is not limited to neural fields, where the framework can be applied modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others\cite{Wikle2002,Xu2005,Fort2008}}. \ken{meteorology [76, 88, 94, 222, 232], epidemiology [123, 137, 218], physics [32, 85, 129, 231], environmental science [45, 87, 183, 204] and economics [31, 48, 165]. (This is copy and paste from my thesis ... parham I'll get you the references later)}. \parham{Mike,Ken: can add to this list?}\mike{While I love this kind of thing for a review, I'm not sure we have space for this kind of thing? Also it's not terribly specific. I think it's better to stick to how it has the potential to model brains in a manner that rocks, rather than loosely saying it could be applied to any spatiotemporal system, i.e. literally everything.}

\mike{The presented framework makes two key assumptions about the cortex: a linear activiation function and stationary dynamics. Our main aim for this framework is to relax these assumptions using the multi-resolution decomposition. While the decomposition itself holds, efficiently performing nonlinear smoothing in the much larger state space remains a challenge. Additional avenues for future work include extending the model to capture a second order synaptic response kernel, and time delays at lower spatial resolutions.  The estimation techniques should also be extended to deal with spatial and temporal heterogeneity in the kernel. Finally, the majority of our current work involves applying this framework to collected data, primarily for the use of seizure detection and mitigation.}

\cut{In order to apply the framework to real data some assumptions must be made, such as stationarity of the parameters over the estimation period. Perhaps the most critical assumption is that the model provides an apt description of the cortical dynamics. The authors acknowledge that there is, and will always be, a discrepancy between the model and cortex. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The key development in this paper is the multi-resolution decomposition forming the state-space model. The decomposition holds for more sophisticated nonlinear forms of the field equations, where the unscented RTS smoother should be used in the E-step of the estimation algorithm.}

\cut{Future work should be directed towards extending the framework to account for a second-order synaptic response kernel, heterogeneity in the connectivity and perhaps time delays at lower spatial resolutions. Most importantly, future work should be directed towards applying and validating the framework on real data.}

>>>>

\usepackage{color}
\newcommand{\mike}[1]{\textcolor{red}{#1}}
\newcommand{\dean}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}} 
\newcommand{\cut}[1]{\textcolor{cyan}{#1}}

@@

\begin{abstract}
%\boldmath
The Integro-difference equation (IDE) is an increasingly popular model of spatio-temporal processes. Here we develop a multi-resolution approximation (MRA) framework for the IDE neural field equations based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximisation (EM) algorithm. A synthetic example is provied
to demonstrate the framework.
\end{abstract}
% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% Neural field model, multiresolution approximation (MRA), Expectation Maximization (EM) algorithm, wavelets.
% \end{IEEEkeywords}
% \IEEEpeerreviewmaketitle
\section{Introduction}
\IEEEPARstart{T}{he} human cerebral cortex has a multi-resolution architecture, where spatial scales for information processing range from ion channels to networks of millions of neurons. This multi-resolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multi-resolution data-driven framework for neural field modeling, the Multi-Resolution Approximate Integro-Difference Equation (MRAIDE), to address this challenge.

@@

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. The current paper is an extension to this recent study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behaviour of the system can be represented simultaneously.

@@

Each $1\times n_{\theta}$ block of the $n_x \times n_x n_{\theta}$ block matrix $\mathbf U$ is

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1~s of data were generated, where the estimation was applied to the final 900~ms, allowing the model's initial transients to die out. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  %\ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} 
The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, \parham{This was obtained by averaging the power (over time \dean{realization?}) of the spatial frequency from the observations} \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). \dean{No assumptions where made about the shape of the kernel.} \parham{It was assumed that the support of the kernel is unknown to the estimator.} The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 1.93$. (b) $j=1$, $n_x=33$, $RMSE = 1.39$. (c) $j=2$, $n_x=65$, $RMSE = 0.90$. (d) $j=3$, $n_x=131$, $RMSE = 0.78$. (e) $j=4$, $n_x=263$, $RMSE=0.78$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig2.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The mean kernel estimates over 200 realizations and \mike{confidence interval} are shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution FEM representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

\cut{The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others\cite{Wikle2002,Xu2005,Fort2008}}. 
% \ken{meteorology [76, 88, 94, 222, 232], epidemiology [123, 137, 218], physics [32, 85, 129, 231], environmental science [45, 87, 183, 204] and economics [31, 48, 165]. (This is copy and paste from my thesis ... parham I'll get you the references later)}. \parham{Mike,Ken: can add to this list?}\mike{While I love this kind of thing for a review, I'm not sure we have space for this kind of thing? Also it's not terribly specific. I think it's better to stick to how it has the potential to model brains in a manner that rocks, rather than loosely saying it could be applied to any spatiotemporal system, i.e. literally everything.}

\mike{The framework makes two key assumptions about the cortex: a linear activation function and stationary dynamics. Our main aim for this framework is to relax these assumptions using the multi-resolution decomposition. \dean{(our main wasn't really to relax the assumptions)} While the decomposition itself holds, efficiently performing nonlinear smoothing in the much larger state space remains a challenge. Additional avenues for future work include extending the model to capture a second order synaptic response kernel, and time delays at lower spatial resolutions.  The estimation techniques should also be extended to deal with spatial and temporal heterogeneity in the kernel. Finally, the majority of our current work involves applying this framework to collected data, primarily for the use of seizure detection and mitigation.}

% \cut{In order to apply the framework to real data some assumptions must be made, such as stationarity of the parameters over the estimation period. Perhaps the most critical assumption is that the model provides an apt description of the cortical dynamics. The authors acknowledge that there is, and will always be, a discrepancy between the model and cortex. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The key development in this paper is the multi-resolution decomposition forming the state-space model. The decomposition holds for more sophisticated nonlinear forms of the field equations, where the unscented RTS smoother should be used in the E-step of the estimation algorithm.}

% \cut{Future work should be directed towards extending the framework to account for a second-order synaptic response kernel, heterogeneity in the connectivity and perhaps time delays at lower spatial resolutions. Most importantly, future work should be directed towards applying and validating the framework on real data.}
